---
title: "Class09 - Candy Project"
author: "Kyle Wittkop (A18592410)"
format: pdf
table-of-contents: true
---

## Importing candy data

```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```


```{r}
nrow(candy)
sum(candy$fruity)
```


> Q1. How many different candy types are in this dataset?

85

> Q2. How many fruity candy types are in the dataset?

38 

```{r}

library(dplyr)

candy %>%
  filter(row.names(candy)=="Twix") %>%
  select(winpercent)

candy["Twix", ]$winpercent

```

>Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?

```{r}
candy["Air Heads", ]$winpercent
```
Air Head win percent = 52.34146

>Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```
76.7686

>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```
49.6535

```{r}
library("skimr")
skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Yes, Win percent seems to be on a different scale, all the other values are between 0 and 1 but winpercent is much larger. 

> Q7. What do you think a zero and one represent for the candy$chocolate column?

```{r}
candy$chocolate
table(candy$chocolate)

```
I think 0 means the candy doesnt contain chocolate and 1 the candy does contain chocolate. 

## Exploratory analysis

> Q8. Plot a histogram of winpercent values

Base R: 

```{r}
hist(candy$winpercent)
```


ggplot: 
```{r}
library(ggplot2)
ggplot(candy,aes(winpercent)) + geom_histogram(bins=30)
```


> Q9. Is the distribution of winpercent values symmetrical?

No, the distribution seems to be right skewed. 

> Q10. Is the center of the distribution above or below 50%?

```{r}
median(candy$winpercent)


```
median of winpercent = 47.82975
seems to be below 50% 

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
chocwin<-candy$winpercent[as.logical(candy$chocolate)]
fruitwin<-candy$winpercent[as.logical(candy$fruity)]

mean(chocwin)
mean(fruitwin)
```
mean of chocolate win percent = 60.92153
mean of fruity win percent = 44.11974 

on average chocolate is higher ranked than fruity. 


> Q12. Is this difference statistically significant?

```{r}
t.test(chocwin,fruitwin)
```
p value = 2.871e-08 which is less than 0.05 so it is considered significant. 

## Overall Candy Rankings


> Q13. What are the five least liked candy types in this set?

```{r}
candy %>% arrange(winpercent) %>% head(5)
```
the 5 least liked candy types are 
1) Nik L Nip
2) Boston Bakes beans
3) Chiclets
4) Super Bubble
5) Jawbusters 


> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy %>% arrange(desc(winpercent)) %>% head(5)
```
The top 5 all time favorite candy types are 
1) Reese's Peanut Butter cup 
2) Reese's Miniatures
3) Twix
4) Kit Kat
5) Snickers 

>Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy) + 
  aes(winpercent, rownames(candy),winpercent) +
  geom_col()
```

```{r}

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col()
```
```{r}
my_cols=rep("green", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "brown"
my_cols[as.logical(candy$bar)] = "blue"
my_cols[as.logical(candy$fruity)] = "red"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

> Q17. What is the worst ranked chocolate candy?

Sixlets 

> Q18. What is the best ranked fruity candy?

Starbursts 

```{r}
library(ggrepel)

ggplot(candy) +
  aes(winpercent,pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 7)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's Miniatures 

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

Nik L Nip 

```{r}

ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_col()
```
```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```

## 6 Exploring the coorelation structure 

```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```
## Principal Component Analysis

>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and Fruity 


>Q23. Similarly, what two variables are most positively correlated?

Chocolate and win percent 



```{r}
pca <- prcomp(candy,scale=TRUE)


summary(pca)

pca$rotation[,1]
```


```{r}
plot(pca$x[,1:2])

```
```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}

my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```
```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

> Q24. Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?

```{r}
ggplot(pca$rotation) +
  aes(PC1, reorder(colnames(candy), PC1)) +
  geom_col()
```
In the positive direction fruity, Sugar percent, Crispedricewafer are seen in the positive direction this makes sense as we saw this previously in the scatter plot. This makes sense to be as we can see choclate, bar and other variables that were negatively cooreleated to fruity in the negative direction. 

## Summary 

> Q25. Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?

The combination of characteristics that make something appear to be a "winning candy" would be that they are chocolate, they are a bar, and have a high price percent. The bar plot of candy ranking based on win percent values showed that bar and chocolate candy were clearly stronger than Fruity and others. price percent vs Win percent graph showed that Bar and chocolate items were both winners and more highly priced.the PCA graphs showed us the positive correlation between Chocolate, Bar, Win percent and pricepercent, which were negatively correlated with fruity, pluribus and hard. 



## Optional extension questions

> Q26. Are popular candies more expensive? In other words: is price significantly different between “winners” and “losers”? List both average values and a P-value along with your answer.

```{r}
losers = candy[which(candy$winpercent < 50),]

winners = candy[which(candy$winpercent >= 50),]

#losers %>% arrange(pricepercent)

#winners %>% arrange(pricepercent)

t.test(winners,losers)

```

Yes, price significantly effects the difference between winners and losers. 



