---
title: "Class12"
author: "Kyle Wittkop (A18592410)"
format: pdf
---

## Section 4: Population Scale Analysis

How many Samples do we have? 

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

```{r}
table(expr$geno)
```
> Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes

```{r}
library(ggplot2)

```
Lets make a boxplot 
```{r}
ggplot(expr) + aes(geno, exp, fill=geno) + 
  geom_boxplot(notch=TRUE)
```

>Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

As seen in the box plot, individuals with the A/A genotype have the highest median expression, while those with the G/G genotype have the lowest median expression. This shows a trend where the presence of the A allele is associated with increased ORMDL3 expression, while the G allele is associated with lower expression. From this pattern, we can infer that the SNP does affect the expression of ORMDL3.