---
title: "Class 5 : Data viz with ggplot"
author: "Kyle Wittkop (PID: A18592410)"
format: pdf
toc: true 
---

## Backround

There are lot's of ways to make figures in R. These include so-called "Base R" graphics (e.g. `plot()`)
and tones of add-on packages like **ggplot2**.

For example here we make the same plot with both: 

```{r}
head(cars)
```

```{r}
plot(cars)
```
First I need to install the package with the command 'install.packages()' 

> **N.B.** we never run an install cmd in a quarto code chunk or we will end up re-installing packages many many times - which is not what we want. 

Everytime we want to use one of these "add-on" packages we need to load it up in R with the `library()` function : 

```{r}
#ggplot(cars)
library(ggplot2)
```
```{r}
ggplot(cars)
```

Every ggplot needs at least 3 things : 

- The **data**, the stuff you want plotted
- the **aes**thetics, how the data will map to the plot
- the **geom**etry, the type of plot 

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) + 
  geom_point() 
```

Add a line to better show relationship between speed and distance 
```{r}
P <- ggplot(cars) + 
  aes(x=speed, y=dist) + 
  geom_point() + geom_smooth(method="lm", se=FALSE,) + 
  labs(title="stopping distance of old cars",
       subtitle="Daa from the `cars` object", 
       x= "Speed (mph)",
       y= "distance (ft)")
```
render it out 
```{r}
P
```

```{r}
P + theme_bw()
```


## Gene Expression Plot 

We can read the input data from the class website 

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

A first version plot 
```{r}
ggplot(genes) + 
  aes(Condition1, Condition2) + 
  geom_point()
```

```{r}
table (genes$State) 
```

Version 2 let's color by 'State' so we can see the up and down signifigant genes copares to all of the unchanging genes 

```{r}
ggplot(genes) + 
  aes(Condition1, Condition2, col=State) + 
  geom_point()
```
Version 3 plot, lets modify the defualt colors to something we like 

```{r}
ggplot(genes) + 
  aes(Condition1, Condition2, col=State) + 
  geom_point() + 
  scale_color_manual(values=c("purple","lightgrey","salmon")) +
  labs(x="Control (no drug)",
       y="Drug", 
        title = "Gene Expression Changes upon GLP-1 drug") + 
  theme_bw()
 
```

## Going further 

lets have a look at the famous 

```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

```{r}
head(gapminder, 6)
```

```{r}
ggplot(gapminder) +
  aes(gdpPercap,lifeExp, col=continent) + 
  geom_point(alpha=0.3) 
```

Lets "facet" (i.e. make a seperate plot) by continent other than the big hot mess above 

```{r}
ggplot(gapminder) +
  aes(gdpPercap,lifeExp, col=continent) + 
  geom_point(alpha=0.3) + 
  facet_wrap(~continent) 
```

##Custom plots
How big is this gapminder dataset ()

```{r}
nrow(gapminder)
library(dplyr)

```

we want to "filter" down to a subset of this data. I will use the **dplyr** package to help me 

First I need to install it and then load it up...
`install.packages("dplyr")` and then `library(dplyr)` 

```{r}
library(dplyr)

 GM2007<- filter(gapminder, year==2007)
 head(GM2007)
 
 
```


```{r}
filter(gapminder, year==2007, country=="Ireland")
```

```{r}
filter(gapminder, year==1977, country=="United States")
```

>Q. Make a plot comparing 1977 and 2007 for all countries 

```{r}
comp<-filter(gapminder, year %in% c(1977,2007))

```
```{r}
ggplot(gapminder) +
  aes(gdpPercap,lifeExp, col=continent) + 
  geom_point(alpha=1) + 
  facet_wrap(~year) + 
  filter(gapminder, year %in% c(1977,2007))
```

