---
title: "Class18"
author: "Kyle Wittkop (A18592510)"
format: pdf
table-of-contents: true
---

## Pertussis and the CMI-PB project

## Backround 
Pertussis (more commonly known as whooping cough) is a highly contagious respiratory disease caused by the bacterium Bordetella pertussis (Figure 1). People of all ages can be infected leading to violent coughing fits followed by a characteristic high-pitched “whoop” like intake of breath. 

## 1. Investigating pertussis cases by year

> Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.


```{r, echo=FALSE}
cdc <- data.frame(
                          Year = c(1922L,
                                   1923L,1924L,1925L,1926L,1927L,1928L,
                                   1929L,1930L,1931L,1932L,1933L,1934L,1935L,
                                   1936L,1937L,1938L,1939L,1940L,1941L,
                                   1942L,1943L,1944L,1945L,1946L,1947L,1948L,
                                   1949L,1950L,1951L,1952L,1953L,1954L,
                                   1955L,1956L,1957L,1958L,1959L,1960L,
                                   1961L,1962L,1963L,1964L,1965L,1966L,1967L,
                                   1968L,1969L,1970L,1971L,1972L,1973L,
                                   1974L,1975L,1976L,1977L,1978L,1979L,1980L,
                                   1981L,1982L,1983L,1984L,1985L,1986L,
                                   1987L,1988L,1989L,1990L,1991L,1992L,1993L,
                                   1994L,1995L,1996L,1997L,1998L,1999L,
                                   2000L,2001L,2002L,2003L,2004L,2005L,
                                   2006L,2007L,2008L,2009L,2010L,2011L,2012L,
                                   2013L,2014L,2015L,2016L,2017L,2018L,
                                   2019L,2020L,2021L,2022L,2023L,2024L,2025L),
  No.Reported.Pertussis.Cases = c(107473,
                                   164191,165418,152003,202210,181411,
                                   161799,197371,166914,172559,215343,179135,
                                   265269,180518,147237,214652,227319,103188,
                                   183866,222202,191383,191890,109873,
                                   133792,109860,156517,74715,69479,120718,
                                   68687,45030,37129,60886,62786,31732,28295,
                                   32148,40005,14809,11468,17749,17135,
                                   13005,6799,7717,9718,4810,3285,4249,
                                   3036,3287,1759,2402,1738,1010,2177,2063,
                                   1623,1730,1248,1895,2463,2276,3589,
                                   4195,2823,3450,4157,4570,2719,4083,6586,
                                   4617,5137,7796,6564,7405,7298,7867,
                                   7580,9771,11647,25827,25616,15632,10454,
                                   13278,16858,27550,18719,48277,28639,
                                   32971,20762,17972,18975,15609,18617,6124,
                                   2116,3044,7063,22538,21906)
)
```

I want a plot of year on the x axis and cases on the y axis 
```{r}
library(ggplot2)

ggplot(cdc) +
  aes(Year,No.Reported.Pertussis.Cases) +
  geom_point() +
  geom_line() + 
  labs(x="Year", y= "Number of cases", title = "Pertussis Cases by Year (1922-2023)")

```

## 2. A tale of two vaccines (wP & aP)

> Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
ggplot(cdc) +
  aes(Year,No.Reported.Pertussis.Cases) +
  geom_point() +
  geom_line() + 
  geom_vline(xintercept=1946, col="blue", lty=2) + 
  geom_vline(xintercept=1996,col="red", lty=2) + 
  geom_vline(xintercept = 2020, col="gray", lty=2)
  labs(x="Year", y= "Number of cases", title = "Pertussis Cases by Year (1922-2023)")

```


> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

aP is shown in plot as the red verticle line. After the introduction of aP, there is an initial period where infections levels seem to remain the same. After around 2022 we can see that there was a large resurgence in the number of cases. A possible explanation of this could be bacteria evolving to evade immunization against the vaccine or it could be caused by less people getting vaccinated causing the resurgence. 

the aP vaccine shows waning efficacy where it reduces immunization for about a 10 year period before the efficacy of the vaccine falls off. 

## 3. Exploring CMI-PB data

```{r}
library(jsonlite)
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 
head(subject, 3)
```
>Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```
there are 87 qP infancy vaccination subjects
there are 85 wP infancy vaccination subjects

> Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```
There are 112 Females and 60 Males 

> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?


```{r}
table(subject$race,subject$biological_sex)
```

32 Asian females, 12 Asian males
2 African American Females, 3 African American Males 
1 Alaska Native
15 Females apart of more than one race, 4 males apart of more than one race
1 Pacific islander female, 1 pacific islander male 
14 unknown females, 7 unknown males
48 white females, 32 white males 

>Q. is this representative of the US population? 

No, there is not enough diversity in the sample to be representative of the US population. 

## Side-Note: Working with dates

```{r}
library(lubridate)
# What is today’s date
today()

# How many days have passed since new year 2000
today() - ymd("2000-01-01")

# What is this in years?
time_length( today() - ymd("2000-01-01"),  "years")
```


> Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}
# Use todays date to calculate age in days
subject$age <- today() - ymd(subject$year_of_birth)
```

(i) the average age of wP 
```{r}
library(dplyr)
wp <- subject %>% filter(infancy_vac == "wP")

round( summary( time_length( wp$age, "years" ) ) )
```
The average Age of wP individuals is 37 years old 

(ii)the average age of aP individuals
```{r}

ap <- subject %>% filter(infancy_vac == "aP")

round( summary( time_length( ap$age, "years" ) ) )

```

The average Age of aP individuals is 28 years old 

(iii)  are they significantly different?
```{r}
x <- t.test(time_length( wp$age, "years" ),
       time_length( ap$age, "years" ))

x$p.value

```
p value < 0.05 

both p values are less than 0.05 meaning the two groups are stastically signifigant 

>Q8. Determine the age of all individuals at time of boost?

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```


>Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")
```
based on the histogram we can clearly see the groups are different 


## Joining multiple tables

```{r}
Specimen <- read_json("http://cmi-pb.org/api/v5_1/specimen", simplifyVector = T)
ab_titer <- read_json("http://cmi-pb.org/api/v5_1/plasma_ab_titer", simplifyVector = T)
```

```{r}
head(Specimen)
```

```{r}
head(ab_titer)
```

To make sense of all this data we need to join (aka merge) all these together, only then will we know that a given aP measurement was collected on a certain data from a certain wP or aP subject 

we can use dplyr and the `*_join()` family of functions to do this 

> Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
meta <- inner_join(Specimen,subject)
head(meta)
```

lets do one more inner join, to join the ab_titer (antibody measurements) with the meta data we just created. 

> Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- inner_join(ab_titer, meta)
head(abdata)
```


>Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```


 IgE   IgG  IgG1  IgG2  IgG3  IgG4 
 6698  7265 11993 12000 12000 12000 
 
lookig at antigens in the data set : 
```{r}
table(abdata$antigen)
```
 
 
>Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(abdata$dataset) 
```
There are 4 number of data set values in abdata. I notice that the number of rows for the most recent data sets is not the largest and is much smaller for the 2020 data sets, suggesting the number of samples collected declined substantially between years.


## 4. Examine IgG Ab titer levels

```{r}
igg <- abdata %>% filter(isotype == "IgG")
head(igg)
```

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() 
```

>Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

FIM2/3, FHA, PT, PRN show the largest difference in level of IgG antibody titers recognized overtime. This is because these are the components of the aP vaccine. 

Others like DT,OVA,TT are not components of the aP vaccine and we wouldnt expect them to show any difference. for example DT  is an antibody for measles, we would not expect the subjects to be producing the antibodies for measles in this data resulting in no antigen difference in level of IgG. 

>Q. is there a difference in these responses between aP and wP individuals? 

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) 
```

Here we can  examine differences between wP and aP here by setting color to include infancy_vac status. 

we can also use faceting : 

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  facet_wrap(~infancy_vac)
```

Is there a difference with time ( a 4 booster shot vs after booster shot)

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac ) +
  geom_boxplot() + 
  facet_wrap(vars(visit))

```


> Q15. Filter to pull out only two specific antigens for analysis and create a boxplot for each. You can chose any you like. Below I picked a “control” antigen (“OVA”, that is not in our vaccines) and a clear antigen of interest (“PT”, Pertussis Toxin, one of the key virulence factors produced by the bacterium B. pertussis).


```{r}
abdata.21 <- abdata %>% filter(dataset == "2021_dataset")

abdata.21 %>% 
  filter(isotype == "IgG", antigen == "PT") %>%
  ggplot(aes(x = planned_day_relative_to_boost,
             y = MFI_normalised,
             col = infancy_vac)) +
  geom_point(aes(group = subject_id), alpha = 0.5) +
  geom_line(aes(group = subject_id), alpha = 0.3) +
  stat_summary(aes(group = infancy_vac),
               fun = mean,
               geom = "line",
               linewidth = 2.5) +
  stat_summary(aes(group = infancy_vac),
               fun = mean,
               geom = "point",
               size = 3) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 14, linetype = "dashed") +
  labs(title = "2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")

```

we can see here the first ecidence that something is different from wP and aP individuals, the peak values are higher in wP and decline slower. 


